---
title: "02. DAR mutation depletion and H/I classification"
output: html_document
---

If a tumour arose from a hepatocyte-like cell, mutations are expected to be
relatively depleted in hepatocyte-open DARs; the converse applies to
biliary-open DARs. This notebook tests that direction without selecting bins
from the same eight labelled samples.

# Inputs and DAR reference

The screened SNV table comes from notebook 04. The two BED files are the
human-coordinate DAR sets produced upstream; overlapping intervals are reduced
before their total lengths are used as denominators.

```{r setup, message=FALSE, warning=FALSE}
# Load genomic-range, table, and plotting packages used in this comparison.
library(GenomicRanges)
library(rtracklayer)
library(dplyr)
library(tidyr)
library(ggplot2)

project_dir <- "C:/Bioinfo data/cell of origin"
out_dir <- file.path(project_dir, "practice_results")
# Read the screened tumour SNVs produced by notebook 04.
snvs <- read.delim(file.path(out_dir, "screened_somatic_snvs.tsv"))

# Restrict overlap calculations to standard human chromosomes.
autosomes_x <- paste0("chr", c(1:22, "X"))
read_dar <- function(path) {
  # Import BED intervals, keep standard chromosomes, and merge overlaps.
  x <- import(path, format = "BED")
  keepSeqlevels(x, intersect(seqlevels(x), autosomes_x),
                pruning.mode = "coarse") %>% reduce()
}
hep_dar <- read_dar("C:/Bioinfo data/hepatocytehuman.bed")
bil_dar <- read_dar("C:/Bioinfo data/billiaryhuman.bed")
hep_mb <- sum(width(hep_dar)) / 1e6
bil_mb <- sum(width(bil_dar)) / 1e6

write.table(
  data.frame(dar_set = c("hepatocyte", "biliary"), Mb = c(hep_mb, bil_mb)),
  file.path(out_dir, "dar_reference_sizes.tsv"), sep = "\t",
  row.names = FALSE, quote = FALSE
)
```

## TMB-scaled DAR counts

The two `per_10000_mutations` columns are the old idea expressed directly:
each sample's DAR count is scaled by that sample's total screened mutation
burden. The lineage score additionally corrects for the different total sizes
of the two DAR sets.

The overlap flags are calculated per SNV, then collapsed to one row per tumour.
The `0.5` pseudocount in `lineage_score` keeps the log-ratio finite when a DAR
set contains no retained mutation.

```{r dar-score}
# Represent each SNV as a one-base genomic range for interval overlap testing.
snv_gr <- makeGRangesFromDataFrame(
  snvs, seqnames.field = "chr", start.field = "pos", end.field = "pos",
  keep.extra.columns = TRUE
)
# Mark whether each SNV falls in a hepatocyte- or biliary-associated DAR.
mcols(snv_gr)$in_hep <- countOverlaps(snv_gr, hep_dar) > 0
mcols(snv_gr)$in_bil <- countOverlaps(snv_gr, bil_dar) > 0

# Collapse the overlap flags to one row per tumour sample.
dar_summary <- as.data.frame(snv_gr) %>%
  group_by(sample) %>%
  summarise(
    total_screened_snvs = n(),
    hepatocyte_dar_snvs = sum(in_hep),
    biliary_dar_snvs = sum(in_bil),
    hepatocyte_per_10000_mutations = 1e4 * hepatocyte_dar_snvs / total_screened_snvs,
    biliary_per_10000_mutations = 1e4 * biliary_dar_snvs / total_screened_snvs,
    hepatocyte_dar_rate_perMb = hepatocyte_dar_snvs / hep_mb,
    biliary_dar_rate_perMb = biliary_dar_snvs / bil_mb,
    # Lower scores are expected for hepatocyte-like samples under the
    # open-chromatin mutation-depletion hypothesis.
    lineage_score = log2(((hepatocyte_dar_snvs + 0.5) / hep_mb) /
                         ((biliary_dar_snvs + 0.5) / bil_mb)),
    .groups = "drop"
  ) %>%
  mutate(
    patient = substr(sample, 1, 5),
    known_label = case_when(
      grepl("^Com0[1-4]H$", sample) ~ "H",
      grepl("^Com0[1-4]I$", sample) ~ "I",
      TRUE ~ NA_character_
    )
  ) %>% arrange(sample)

write.table(dar_summary, file.path(out_dir, "dar_enrichment_summary.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
dar_summary
```

## Paired test and leave-one-patient-out validation

Lower score predicts H; higher score predicts I. Each held-out patient's cutoff
is estimated only from the other three patients.

The paired sign test asks only whether the direction is consistent across
patients; it does not assume normally distributed scores. LOPO then checks
whether the same rule generalises to a patient excluded from cutoff estimation.

```{r validation}
# Keep the eight labelled H/I samples for paired testing and validation.
known <- filter(dar_summary, !is.na(known_label))
# Use one classifier rule for both LOPO folds and any future unlabelled sample.
classify_score <- function(score, cutoff, direction_consistent) {
  if (direction_consistent) if_else(score < cutoff, "H", "I")
  else rep("Indeterminate", length(score))
}

# Compare H and I scores within each patient rather than across patients.
paired <- known %>%
  select(patient, known_label, lineage_score) %>%
  pivot_wider(names_from = known_label, values_from = lineage_score) %>%
  mutate(I_minus_H = I - H,
         direction = if_else(I_minus_H > 0, "expected", "opposite"))

# Test whether the expected I-minus-H direction is consistent across patients.
sign_test <- binom.test(sum(paired$I_minus_H > 0), nrow(paired), p = 0.5)

lopo <- lapply(unique(known$patient), function(held_out) {
  # Estimate the midpoint using the other patients only.
  train <- filter(known, patient != held_out)
  test <- filter(known, patient == held_out)
  mean_h <- mean(train$lineage_score[train$known_label == "H"])
  mean_i <- mean(train$lineage_score[train$known_label == "I"])
  cutoff <- (mean_h + mean_i) / 2
  # Do not classify a fold whose training data reverse the prespecified direction.
  direction_consistent <- mean_h < mean_i
  mutate(test, held_patient = held_out, fold_cutoff = cutoff,
         predicted = classify_score(lineage_score, cutoff, direction_consistent),
         evaluated = predicted != "Indeterminate",
         correct = evaluated & predicted == known_label,
         training_direction_consistent = direction_consistent)
}) %>% bind_rows()

n_evaluated <- sum(lopo$evaluated)
lopo_accuracy <- if (n_evaluated) mean(lopo$correct[lopo$evaluated]) else NA_real_

write.table(paired, file.path(out_dir, "paired_dar_score_differences.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
write.table(lopo, file.path(out_dir, "leave_one_patient_out_predictions.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)

paired
sign_test
mean(lopo$correct)
```

## Score plot

Each line connects the H and I samples from one patient. A consistent expected
pattern would show most lines rising from H to I.

```{r plot, fig.width=6, fig.height=4}
# Connect each patient's H and I scores to show the paired direction.
ggplot(known, aes(known_label, lineage_score, group = patient, colour = patient)) +
  geom_line() + geom_point(size = 3) +
  labs(x = NULL, y = "log2(H-DAR / I-DAR mutation density)") +
  theme_classic()
```

## Final summary

```{r mix02-and-result}
# Summarise the paired test and held-out predictions for the final report.
result <- c(
  sprintf("Expected paired direction: %d/%d", sum(paired$I_minus_H > 0), nrow(paired)),
  sprintf("Paired sign-test p = %.3g", sign_test$p.value),
  sprintf("LOPO accuracy: %d/%d evaluated (%.1f%%); %d indeterminate",
          sum(lopo$correct), n_evaluated, 100 * lopo_accuracy,
          sum(!lopo$evaluated)),
  "Conclusion: no reliable H/I classifier is supported by these data."
)
writeLines(result, file.path(out_dir, "final_result.txt"))
cat(paste(result, collapse = "\n"))
```

Main limitations: only four paired patients, candidate-like Mutect2 WES calls,
no capture/callable BED, and mouse-to-human lifted DARs.
