---
title: "Top 10% Bin DAR Mutation Analysis"
author:
- name: Zack Guo
  affiliation: UCL
  email: wegmzgu@ucl.ac.uk
date: "`r format(Sys.time(), '%B %e, %Y')`"
output:
  html_document:
    highlight: tango
    toc: yes
    toc_float: yes
    toc_depth: '5'
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Library packages

```{r}
# Load packages for genomic overlaps, tables, and plotting.
library(GenomicRanges)
library(GenomeInfoDb)
library(rtracklayer)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Aim

This notebook selects the 10% of 1 Mb genomic bins with the largest scaled
mutation-count difference between the eight labelled H and I tumour samples.
The hepatocyte- and biliary-associated DARs are restricted to those bins, and
the original DAR mutation score is then calculated for the eight labelled
samples.

# Input data

Notebook 04 screens the eight `Com01-04` H/I tumour-normal VCFs. These same
eight samples are used for the top-bin selection and paired comparison.

```{r inputs}
project_dir <- "C:/Bioinfo data/cell of origin"
out_dir <- file.path(project_dir, "practice_results")

# Read the eight-sample screened SNV table produced by notebook 04.
input_path <- file.path(out_dir, "screened_somatic_snvs.tsv")
stopifnot(file.exists(input_path))
snvs <- read.delim(input_path)

labelled_samples <- as.vector(rbind(
  paste0("Com0", 1:4, "H"),
  paste0("Com0", 1:4, "I")
))
# Ignore any stale extra sample rows if notebook 05 is run before notebook 04.
snvs <- filter(snvs, sample %in% labelled_samples)
stopifnot(setequal(unique(snvs$sample), labelled_samples))

data.frame(
  sample = labelled_samples,
  screened_snvs = as.integer(table(snvs$sample)[labelled_samples]),
  source_file = basename(input_path)
)
```

# Create and rank 1 Mb bins

The hg19 autosomes and chromosome X are divided into 3,053 bins. Within each
labelled sample, the mutation count in each bin is divided by that sample's
total screened SNV count and multiplied by 10,000.

For each bin, the four H values and four I values are averaged separately. Bins
are ranked by `abs(mean_I - mean_H)`, and the highest-ranking 10% are retained:
`ceiling(3053 * 0.10) = 306` bins.

```{r select-top-bins}
# Define hg19 chromosome lengths for chromosomes 1-22 and X.
hg19_lengths <- c(
  chr1 = 249250621, chr2 = 243199373, chr3 = 198022430,
  chr4 = 191154276, chr5 = 180915260, chr6 = 171115067,
  chr7 = 159138663, chr8 = 146364022, chr9 = 141213431,
  chr10 = 135534747, chr11 = 135006516, chr12 = 133851895,
  chr13 = 115169878, chr14 = 107349540, chr15 = 102531392,
  chr16 = 90354753, chr17 = 81195210, chr18 = 78077248,
  chr19 = 59128983, chr20 = 63025520, chr21 = 48129895,
  chr22 = 51304566, chrX = 155270560
)

# Create all 1 Mb bins, including bins with zero mutations.
genome_bins <- tileGenome(
  hg19_lengths,
  tilewidth = 1e6,
  cut.last.tile.in.chrom = TRUE
)
names(genome_bins) <- sprintf("bin_%04d", seq_along(genome_bins))

total_bins <- length(genome_bins)
n_top_bins <- ceiling(total_bins * 0.10)

# Convert the screened SNVs to one-base genomic ranges.
snv_gr <- makeGRangesFromDataFrame(
  snvs,
  seqnames.field = "chr",
  start.field = "pos",
  end.field = "pos",
  keep.extra.columns = TRUE
)

# Count and scale mutations in every bin for each labelled sample.
bin_counts <- lapply(labelled_samples, function(sample_id) {
  sample_gr <- snv_gr[mcols(snv_gr)$sample == sample_id]
  mutation_count <- countOverlaps(genome_bins, sample_gr)

  data.frame(
    sample = sample_id,
    bin_id = names(genome_bins),
    chr = as.character(seqnames(genome_bins)),
    start = start(genome_bins),
    end = end(genome_bins),
    mutation_count = mutation_count,
    sample_total = length(sample_gr),
    mutations_per_10000_screened =
      1e4 * mutation_count / length(sample_gr)
  )
}) %>% bind_rows()

# Place the eight scaled sample values on one row per bin.
bin_ranking <- bin_counts %>%
  select(bin_id, chr, start, end, sample, mutations_per_10000_screened) %>%
  pivot_wider(
    names_from = sample,
    values_from = mutations_per_10000_screened
  )

# Calculate and rank the absolute H-versus-I group difference.
h_columns <- grep("H$", names(bin_ranking), value = TRUE)
i_columns <- grep("I$", names(bin_ranking), value = TRUE)
bin_ranking$mean_H <- rowMeans(bin_ranking[h_columns])
bin_ranking$mean_I <- rowMeans(bin_ranking[i_columns])
bin_ranking$I_minus_H <- bin_ranking$mean_I - bin_ranking$mean_H
bin_ranking$absolute_difference <- abs(bin_ranking$I_minus_H)

bin_ranking <- bin_ranking %>%
  arrange(desc(absolute_difference), chr, start) %>%
  mutate(difference_rank = row_number())

# Select the 306 highest-ranking bins.
top_bin_ids <- head(bin_ranking$bin_id, n_top_bins)
top_bins <- genome_bins[names(genome_bins) %in% top_bin_ids]

write.table(
  bin_ranking,
  file.path(out_dir, "top10_percent_bin_ranking.tsv"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

data.frame(total_bins = total_bins, selected_bins = length(top_bins))
```

# Restrict DARs to the top 10% bins

The hepatocyte and biliary DARs are imported, reduced, and intersected with the
same 306 bins. Their intersected lengths are used as the mutation-rate
denominators.

```{r top-bin-dars}
autosomes_x <- names(hg19_lengths)

# Import one human-coordinate DAR BED file.
read_dar <- function(path) {
  dar <- import(path, format = "BED")
  keepSeqlevels(
    dar,
    intersect(seqlevels(dar), autosomes_x),
    pruning.mode = "coarse"
  ) %>% reduce()
}

# Intersect an unstranded DAR reference with the selected genomic bins.
restrict_dar <- function(dar, bins) {
  strand(dar) <- "*"
  strand(bins) <- "*"
  reduce(GenomicRanges::intersect(dar, bins))
}

hep_dar_top <- restrict_dar(
  read_dar("C:/Bioinfo data/hepatocytehuman.bed"),
  top_bins
)
bil_dar_top <- restrict_dar(
  read_dar("C:/Bioinfo data/billiaryhuman.bed"),
  top_bins
)

hep_mb <- sum(width(hep_dar_top)) / 1e6
bil_mb <- sum(width(bil_dar_top)) / 1e6
stopifnot(hep_mb > 0, bil_mb > 0)

dar_reference_summary <- data.frame(
  dar_set = c("hepatocyte", "biliary"),
  selected_bins = n_top_bins,
  Mb = c(hep_mb, bil_mb)
)

write.table(
  dar_reference_summary,
  file.path(out_dir, "top10_percent_dar_reference_sizes.tsv"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

dar_reference_summary
```

# Calculate top-bin DAR scores

DAR mutation counts are scaled by total screened mutation burden, while the
lineage score also corrects for the different total sizes of the two restricted
DAR sets. The `0.5` pseudocount keeps the log-ratio finite when a DAR set
contains no retained mutation.

```{r dar-scores}
# Apply the same top-bin DAR score to each of the eight samples.
dar_summary <- lapply(labelled_samples, function(sample_id) {
  sample_gr <- snv_gr[mcols(snv_gr)$sample == sample_id]
  hep_count <- sum(countOverlaps(sample_gr, hep_dar_top) > 0)
  bil_count <- sum(countOverlaps(sample_gr, bil_dar_top) > 0)

  data.frame(
    sample = sample_id,
    patient = substr(sample_id, 1, 5),
    known_label = case_when(
      grepl("^Com0[1-4]H$", sample_id) ~ "H",
      grepl("^Com0[1-4]I$", sample_id) ~ "I"
    ),
    total_screened_snvs = length(sample_gr),
    hepatocyte_dar_snvs = hep_count,
    biliary_dar_snvs = bil_count,
    hepatocyte_per_10000_mutations =
      1e4 * hep_count / length(sample_gr),
    biliary_per_10000_mutations =
      1e4 * bil_count / length(sample_gr),
    hepatocyte_dar_rate_perMb = hep_count / hep_mb,
    biliary_dar_rate_perMb = bil_count / bil_mb,
    lineage_score = log2(
      ((hep_count + 0.5) / hep_mb) /
        ((bil_count + 0.5) / bil_mb)
    )
  )
}) %>% bind_rows()

write.table(
  dar_summary,
  file.path(out_dir, "top10_percent_dar_enrichment_summary.tsv"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

dar_summary
```

# Paired H/I result

Lower scores are expected for H samples and higher scores for I samples.

```{r paired-result}
# Compare H and I within each patient.
paired <- dar_summary %>%
  select(patient, known_label, lineage_score) %>%
  pivot_wider(names_from = known_label, values_from = lineage_score) %>%
  mutate(
    I_minus_H = I - H,
    direction = if_else(I_minus_H > 0, "expected", "opposite")
  )

sign_test <- binom.test(
  sum(paired$I_minus_H > 0),
  nrow(paired),
  p = 0.5
)

write.table(
  paired,
  file.path(out_dir, "top10_percent_paired_differences.tsv"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

paired
sign_test
```

# Line plot

Each coloured line connects the H and I components from one patient. A
consistent expected pattern has lines rising from H to I.

```{r plot, fig.width=6, fig.height=4}
# Connect each patient's H and I scores to show the paired direction.
ggplot(
  dar_summary,
  aes(known_label, lineage_score, group = patient, colour = patient)
) +
  geom_line() +
  geom_point(size = 3) +
  labs(
    x = NULL,
    y = "Top-10% log2(H-DAR / I-DAR mutation density)",
    colour = "Patient"
  ) +
  theme_classic()
```

# Final summary

```{r final-result}
# Write the paired result as a short text summary.
result <- sprintf(
  "Top 10%% bins: expected paired direction %d/%d; sign-test p = %.3g",
  sum(paired$I_minus_H > 0),
  nrow(paired),
  sign_test$p.value
)

writeLines(result, file.path(out_dir, "final_result.txt"))
cat(paste(result, collapse = "\n"))
```

Main limitations: only four labelled patient pairs, use of the same samples for
bin selection and paired comparison, sparse DAR mutation counts, candidate-like
Mutect2 WES calls, no capture/callable-region BED, and mouse-to-human lifted
DARs.
