---
title: "DAR Mutation Depletion and H/I Comparison"
author:
- name: Zack Guo
  affiliation: UCL
  email: wegmzgu@ucl.ac.uk
date: "`r format(Sys.time(), '%B %e, %Y')`"
output:
  html_document:
    highlight: tango
    toc: yes
    toc_float: yes
    toc_depth: '5'
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Library packages
```{r}
# Load genomic-range, table, and plotting packages used in this comparison.
library(GenomicRanges)
library(rtracklayer)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Aim
If a tumour arose from a hepatocyte-like cell, mutations are expected to be
relatively depleted in hepatocyte-open DARs; the converse applies to
biliary-open DARs. This notebook tests that direction without selecting bins
from the same eight labelled samples.

# Inputs and DAR reference

The screened SNV table comes from notebook 04. The two BED files are the
human-coordinate DAR sets produced upstream; overlapping intervals are reduced
before their total lengths are used as denominators.

```{r}
project_dir <- "C:/Bioinfo data/cell of origin"
out_dir <- file.path(project_dir, "practice_results")
# Read the screened tumour SNVs produced by notebook 04.
snvs <- read.delim(file.path(out_dir, "screened_somatic_snvs.tsv"))

# Restrict overlap calculations to standard human chromosomes.
autosomes_x <- paste0("chr", c(1:22, "X"))
read_dar <- function(path) {
  # Import BED intervals, keep standard chromosomes, and merge overlaps.
  x <- import(path, format = "BED")
  keepSeqlevels(x, intersect(seqlevels(x), autosomes_x),
                pruning.mode = "coarse") %>% reduce()
}
hep_dar <- read_dar("C:/Bioinfo data/hepatocytehuman.bed")
bil_dar <- read_dar("C:/Bioinfo data/billiaryhuman.bed")
hep_mb <- sum(width(hep_dar)) / 1e6
bil_mb <- sum(width(bil_dar)) / 1e6

write.table(
  data.frame(dar_set = c("hepatocyte", "biliary"), Mb = c(hep_mb, bil_mb)),
  file.path(out_dir, "dar_reference_sizes.tsv"), sep = "\t",
  row.names = FALSE, quote = FALSE
)
```

# TMB-scaled DAR counts

The two `per_10000_mutations` columns are the old idea expressed directly:
each sample's DAR count is scaled by that sample's total screened mutation
burden. The lineage score additionally corrects for the different total sizes
of the two DAR sets.

The overlap flags are calculated per SNV, then collapsed to one row per tumour.
The `0.5` pseudocount in `lineage_score` keeps the log-ratio finite when a DAR
set contains no retained mutation.

```{r dar-score}
# Represent each SNV as a one-base genomic range for interval overlap testing.
snv_gr <- makeGRangesFromDataFrame(
  snvs, seqnames.field = "chr", start.field = "pos", end.field = "pos",
  keep.extra.columns = TRUE
)
# Mark whether each SNV falls in a hepatocyte- or biliary-associated DAR.
mcols(snv_gr)$in_hep <- countOverlaps(snv_gr, hep_dar) > 0
mcols(snv_gr)$in_bil <- countOverlaps(snv_gr, bil_dar) > 0

# Collapse the overlap flags to one row per tumour sample.
dar_summary <- as.data.frame(snv_gr) %>%
  group_by(sample) %>%
  summarise(
    total_screened_snvs = n(),
    hepatocyte_dar_snvs = sum(in_hep),
    biliary_dar_snvs = sum(in_bil),
    hepatocyte_per_10000_mutations = 1e4 * hepatocyte_dar_snvs / total_screened_snvs,
    biliary_per_10000_mutations = 1e4 * biliary_dar_snvs / total_screened_snvs,
    hepatocyte_dar_rate_perMb = hepatocyte_dar_snvs / hep_mb,
    biliary_dar_rate_perMb = biliary_dar_snvs / bil_mb,
    # Lower scores are expected for hepatocyte-like samples under the
    # open-chromatin mutation-depletion hypothesis.
    lineage_score = log2(((hepatocyte_dar_snvs + 0.5) / hep_mb) /
                         ((biliary_dar_snvs + 0.5) / bil_mb)),
    .groups = "drop"
  ) %>%
  mutate(
    patient = substr(sample, 1, 5),
    known_label = case_when(
      grepl("^Com0[1-4]H$", sample) ~ "H",
      grepl("^Com0[1-4]I$", sample) ~ "I",
      TRUE ~ NA_character_
    )
  ) %>% arrange(sample)

write.table(dar_summary, file.path(out_dir, "dar_enrichment_summary.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
dar_summary
```

# Paired direction test

Lower score is expected for H samples and higher score for I samples.

The paired sign test asks only whether this direction is consistent across
patients; it is a descriptive consistency check rather than a predictive model.

```{r paired-test}
# Keep the eight labelled H/I samples for the paired direction test.
known <- filter(dar_summary, !is.na(known_label))

# Compare H and I scores within each patient rather than across patients.
paired <- known %>%
  select(patient, known_label, lineage_score) %>%
  pivot_wider(names_from = known_label, values_from = lineage_score) %>%
  mutate(I_minus_H = I - H,
         direction = if_else(I_minus_H > 0, "expected", "opposite"))

# Test whether the expected I-minus-H direction is consistent across patients.
sign_test <- binom.test(sum(paired$I_minus_H > 0), nrow(paired), p = 0.5)

write.table(paired, file.path(out_dir, "paired_dar_score_differences.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)

paired
sign_test
```

# Score plot

Each line connects the H and I samples from one patient. A consistent expected
pattern would show most lines rising from H to I.

```{r plot, fig.width=6, fig.height=4}
# Connect each patient's H and I scores to show the paired direction.
ggplot(known, aes(known_label, lineage_score, group = patient, colour = patient)) +
  geom_line() + geom_point(size = 3) +
  labs(x = NULL, y = "log2(H-DAR / I-DAR mutation density)") +
  theme_classic()
```

# Final summary

```{r final-result}
# Summarise the paired direction test for the final report.
result <- c(
  sprintf("Expected paired direction: %d/%d", sum(paired$I_minus_H > 0), nrow(paired)),
  sprintf("Paired sign-test p = %.3g", sign_test$p.value),
  "Conclusion: the paired direction test does not provide reliable H/I separation."
)
writeLines(result, file.path(out_dir, "final_result.txt"))
cat(paste(result, collapse = "\n"))
```

Main limitations: only four paired patients, candidate-like Mutect2 WES calls,
no capture/callable BED, and mouse-to-human lifted DARs.
