---
title: "scATAC integration"
author:
- name: Zack Guo
  affiliation: UCL
  email: wegmzgu@ucl.ac.uk
date: "`r format(Sys.time(), '%B %e, %Y')`"
output:
  html_document:
    highlight: tango
    toc: yes
    toc_float: yes
    toc_depth: '5'
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Library packages
```{r}
library(Matrix)
library(Seurat)
library(Signac)
library(dplyr)
library(ggplot2)
library(GenomicRanges)
library(GenomeInfoDb)
library(harmony)
library(rtracklayer)
```

# scATAC integration. Parameter and filtering follows the paper where possible with local peak matrix files.
```{r}
setwd("C:/Bioinfo data")

# Define function to read Cell Ranger ATAC peak matrix
load_atac_sample <- function(base_dir, sample_name, sample_label, condition_label) {
  base_dir <- gsub("\\\\", "/", base_dir)
  
  peak_path <- file.path(base_dir, paste0(sample_name, "_peaks.bed.gz"))
  matrix_path <- file.path(base_dir, paste0(sample_name, "_matrix.mtx.gz"))
  barcode_path <- file.path(base_dir, paste0(sample_name, "_barcodes.tsv.gz"))
  
  peaks <- read.table(
    gzfile(peak_path),
    sep = "\t",
    header = FALSE,
    stringsAsFactors = FALSE
  )
  colnames(peaks) <- c("chr", "start", "end")
  
  peak_names <- paste(peaks$chr, peaks$start, peaks$end, sep = "-")
  
  counts <- readMM(gzfile(matrix_path))
  barcodes <- readLines(gzfile(barcode_path))
  
  rownames(counts) <- peak_names
  colnames(counts) <- paste(sample_label, barcodes, sep = "_")
  
  peak_ranges <- makeGRangesFromDataFrame(
    peaks,
    seqnames.field = "chr",
    start.field = "start",
    end.field = "end"
  )
  names(peak_ranges) <- peak_names
  
  chrom_assay <- CreateChromatinAssay(
    counts = counts,
    ranges = peak_ranges,
    genome = "hg19"
  )
  
  seu <- CreateSeuratObject(
    counts = chrom_assay,
    assay = "peaks",
    project = sample_label
  )
  
  seu$sample <- sample_label
  seu$condition <- condition_label
  
  return(seu)
}

# Load four ATAC samples
base_dir <- "C:/Bioinfo data/Used before/ATAC"

dm1 <- load_atac_sample(base_dir, "GSM4832485_DM-1_hg19", "DM_1", "DM")
dm2 <- load_atac_sample(base_dir, "GSM4832486_DM-2_hg19", "DM_2", "DM")
hm1 <- load_atac_sample(base_dir, "GSM4832487_HM-1_hg19", "HM_1", "HM")
hm2 <- load_atac_sample(base_dir, "GSM4832488_HM-2_hg19", "HM_2", "HM")

# Merge four ATAC samples
atac <- merge(
  dm1,
  y = list(dm2, hm1, hm2)
)

# Load hg19 gene annotation from GRCh37 GTF
gtf_path <- "C:/Bioinfo data/Used before/ATAC/annotation/Homo_sapiens.GRCh37.87.gtf.gz"

file.exists(gtf_path)

gene_annot <- rtracklayer::import(gtf_path)

# Keep gene-level annotation only
gene_annot <- gene_annot[gene_annot$type == "gene"]

# Match chromosome naming style to hg19 peaks
seqlevelsStyle(gene_annot) <- "UCSC"

# Keep standard chromosomes only
gene_annot <- keepStandardChromosomes(
  gene_annot,
  pruning.mode = "coarse"
)

# Set genome information
genome(gene_annot) <- "hg19"

# Add gene annotation to ATAC object
Annotation(atac) <- gene_annot
```

# ATAC dimension reduction and batch correction
```{r}
DefaultAssay(atac) <- "peaks"

# TF-IDF and LSI follow the standard scATAC workflow. Dimension 1 is usually sequencing depth.
atac <- FindTopFeatures(atac, min.cutoff = 10)
atac <- RunTFIDF(atac)
atac <- RunSVD(atac, n = 30)

# Harmony batch correction using sample as batch variable
atac <- RunHarmony(
  object = atac,
  group.by.vars = "sample",
  reduction = "lsi",
  dims.use = 1:30,
  reduction.save = "harmony",
  project.dim = FALSE
)

# Use dimensions 2 to 30 for downstream analysis because LSI dimension 1 often captures sequencing depth
atac <- RunUMAP(atac, reduction = "harmony", dims = 2:30)
atac <- FindNeighbors(atac, reduction = "harmony", dims = 2:30)
atac <- FindClusters(atac, resolution = 0.6)

DimPlot(atac, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
```

# Gene activity matrix for scRNA-scATAC integration
```{r}
# Gene activity score is approximated from the peak count matrix because fragment files are not available
DefaultAssay(atac) <- "peaks"

# Get peak count matrix
peak_counts <- GetAssayData(
  atac,
  assay = "peaks",
  layer = "counts"
)

# Get peak genomic ranges
peak_ranges <- granges(atac[["peaks"]])

# Check that peak ranges and peak count matrix are in the same order
stopifnot(length(peak_ranges) == nrow(peak_counts))

# Manually add peak names because granges() may return unnamed GRanges
names(peak_ranges) <- rownames(peak_counts)

# Get gene annotation
gene_ranges <- Annotation(atac)

# Keep gene-level annotation only
gene_ranges <- gene_ranges[gene_ranges$type == "gene"]

# Get Ensembl gene IDs and remove version suffix if present
gene_ids <- as.character(mcols(gene_ranges)$gene_id)
gene_ids <- sub("\\.\\d+$", "", gene_ids)

# Keep genes with valid Ensembl IDs
keep_genes <- !is.na(gene_ids) & gene_ids != ""
gene_ranges <- gene_ranges[keep_genes]
gene_ids <- gene_ids[keep_genes]

# Extend gene body by 2 kb upstream of TSS
gene_windows <- gene_ranges

plus_strand <- as.character(strand(gene_windows)) == "+"
minus_strand <- as.character(strand(gene_windows)) == "-"
unknown_strand <- as.character(strand(gene_windows)) == "*"

start(gene_windows)[plus_strand] <- pmax(
  1,
  start(gene_windows)[plus_strand] - 2000
)

end(gene_windows)[minus_strand] <- end(gene_windows)[minus_strand] + 2000

start(gene_windows)[unknown_strand] <- pmax(
  1,
  start(gene_windows)[unknown_strand] - 2000
)

# Find peak-gene overlaps
overlaps <- findOverlaps(
  query = gene_windows,
  subject = peak_ranges,
  ignore.strand = TRUE
)

# Build gene-by-peak mapping matrix
unique_gene_ids <- unique(gene_ids)

gene_peak_matrix <- sparseMatrix(
  i = match(gene_ids[queryHits(overlaps)], unique_gene_ids),
  j = subjectHits(overlaps),
  x = 1,
  dims = c(length(unique_gene_ids), length(peak_ranges)),
  dimnames = list(unique_gene_ids, rownames(peak_counts))
)

# Make overlap matrix binary
gene_peak_matrix@x <- rep(1, length(gene_peak_matrix@x))

# Sum peak counts for each gene
gene.activities <- gene_peak_matrix %*% peak_counts

# Keep genes with non-zero activity
gene.activities <- gene.activities[
  Matrix::rowSums(gene.activities) > 0,
]

# Add gene activity matrix as a new assay
atac[["ACTIVITY"]] <- CreateAssayObject(
  counts = gene.activities
)

DefaultAssay(atac) <- "ACTIVITY"

atac <- NormalizeData(atac)

atac <- ScaleData(
  atac,
  features = rownames(atac)
)

# Check result
dim(gene.activities)
head(rownames(gene.activities))
```

# Load annotated scRNA reference saved from the clustering Rmd
```{r}
rna_reference <- readRDS("C:/Bioinfo data/RDS to skip calc/HM_DM_scRNA_reference_for_ATAC.rds")

DefaultAssay(rna_reference) <- "RNA"
Idents(rna_reference) <- "cell_type"

table(rna_reference$seurat_clusters)
table(rna_reference$cell_type)
```

# Transfer scRNA cell type labels to scATAC cells
```{r}
DefaultAssay(rna_reference) <- "RNA"
DefaultAssay(atac) <- "ACTIVITY"

# Select variable genes from scRNA reference
rna_reference <- FindVariableFeatures(
  rna_reference,
  selection.method = "vst",
  nfeatures = 1500
)

# Use shared variable genes for label transfer
transfer_features <- intersect(
  VariableFeatures(rna_reference),
  rownames(atac[["ACTIVITY"]])
)

length(transfer_features)

# Scale selected transfer genes
rna_reference <- ScaleData(
  rna_reference,
  assay = "RNA",
  features = transfer_features
)

atac <- ScaleData(
  atac,
  assay = "ACTIVITY",
  features = transfer_features
)

# Find transfer anchors
transfer_anchors <- FindTransferAnchors(
  reference = rna_reference,
  query = atac,
  reference.assay = "RNA",
  query.assay = "ACTIVITY",
  reduction = "cca",
  features = transfer_features,
  dims = 1:15
)

# Transfer scRNA cell type labels to scATAC cells
predictions <- TransferData(
  anchorset = transfer_anchors,
  refdata = rna_reference$cell_type,
  weight.reduction = atac[["harmony"]],
  dims = 2:15
)

atac <- AddMetaData(
  atac,
  metadata = predictions
)

# Remove low-confidence predicted cells
atac_filtered <- subset(
  atac,
  subset = prediction.score.max >= 0.3
)

table(atac$predicted.id)
table(atac_filtered$predicted.id)
table(atac_filtered$condition, atac_filtered$predicted.id)

DimPlot(
  atac_filtered,
  reduction = "umap",
  group.by = "predicted.id",
  label = TRUE,
  repel = TRUE
) + NoLegend()
```

# Save integrated scATAC object for downstream analysis
```{r}
saveRDS(
  list(
    atac = atac,
    atac_filtered = atac_filtered,
    predictions = predictions
  ),
  "C:/Bioinfo data/RDS to skip calc/HM_DM_scATAC_integrated_with_scRNA_labels.rds"
)
```

# Quality check

```{r}
# Load the saved integrated scATAC object before running quality checks.
# This avoids re-running the upstream integration and label transfer steps.
integrated_atac_rds <- "C:/Bioinfo data/RDS to skip calc/HM_DM_scATAC_integrated_with_scRNA_labels.rds"

integrated_atac <- readRDS(integrated_atac_rds)

# Restore the objects saved from the label transfer step.
atac <- integrated_atac$atac
atac_filtered <- integrated_atac$atac_filtered
predictions <- integrated_atac$predictions

# Use gene activity as the default assay for downstream marker checks.
DefaultAssay(atac) <- "ACTIVITY"
DefaultAssay(atac_filtered) <- "ACTIVITY"

# Quick check that the expected objects and dimensions were loaded.
names(integrated_atac)
dim(atac)
dim(atac_filtered)
```

```{r}
# Check 1: summarise label transfer confidence.
# Low or moderate prediction scores suggest weak scRNA-to-scATAC label transfer.
meta_atac <- atac@meta.data

# Overall confidence across all ATAC cells.
overall_confidence <- data.frame(
  total_cells = nrow(meta_atac),
  median_prediction_score = median(meta_atac$prediction.score.max, na.rm = TRUE),
  mean_prediction_score = mean(meta_atac$prediction.score.max, na.rm = TRUE),
  cells_above_0.3 = sum(meta_atac$prediction.score.max >= 0.3, na.rm = TRUE),
  pass_rate_above_0.3 = mean(meta_atac$prediction.score.max >= 0.3, na.rm = TRUE)
)

overall_confidence

# Confidence summary for each transferred cell type.
confidence_by_label <- meta_atac %>%
  group_by(predicted.id) %>%
  summarise(
    n_cells = n(),
    median_prediction_score = median(prediction.score.max, na.rm = TRUE),
    mean_prediction_score = mean(prediction.score.max, na.rm = TRUE),
    pass_rate_above_0.3 = mean(prediction.score.max >= 0.3, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_cells))

confidence_by_label

# Distribution of prediction scores across all cells.
ggplot(meta_atac, aes(x = prediction.score.max)) +
  geom_histogram(bins = 50) +
  theme_classic() +
  labs(
    title = "Distribution of label transfer prediction scores",
    x = "Prediction score",
    y = "Number of ATAC cells"
  )

# Prediction score spread for each transferred label.
ggplot(meta_atac, aes(x = predicted.id, y = prediction.score.max)) +
  geom_boxplot(outlier.size = 0.2) +
  coord_flip() +
  theme_classic() +
  labs(
    title = "Prediction score by transferred cell type",
    x = "Predicted cell type",
    y = "Prediction score"
  )
```

```{r}
# Check 2: compare unsupervised ATAC clusters with transferred scRNA labels.
# Good label transfer should give one dominant label within most ATAC clusters.
cluster_label_counts <- as.data.frame(
  table(atac$seurat_clusters, atac$predicted.id)
)

colnames(cluster_label_counts) <- c(
  "atac_cluster",
  "predicted_label",
  "n_cells"
)

# Convert counts to within-cluster fractions.
cluster_label_props <- cluster_label_counts %>%
  group_by(atac_cluster) %>%
  mutate(
    cluster_total = sum(n_cells),
    fraction = n_cells / cluster_total
  ) %>%
  ungroup()

# For each ATAC cluster, identify the most common transferred label.
cluster_purity <- cluster_label_props %>%
  group_by(atac_cluster) %>%
  slice_max(fraction, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(as.numeric(as.character(atac_cluster)))

cluster_purity

# Heatmap of label composition within each ATAC cluster.
ggplot(
  cluster_label_props,
  aes(x = atac_cluster, y = predicted_label, fill = fraction)
) +
  geom_tile() +
  geom_text(aes(label = round(fraction, 2)), size = 3) +
  theme_classic() +
  labs(
    title = "Consistency between ATAC clusters and transferred labels",
    x = "Unsupervised ATAC cluster",
    y = "Transferred scRNA label",
    fill = "Fraction"
  )
```

```{r}
# Check 3: inspect marker gene activity in transferred cell types.
# Reliable transferred labels should show expected marker activity patterns.
DefaultAssay(atac_filtered) <- "ACTIVITY"

# Markers selected to test hepatocyte, cholangiocyte, progenitor, and stromal identities.
lineage_markers <- c(
  "FGG",
  "ACSM3",
  "APOC1",
  "CYP3A5",
  "TFF1",
  "KRT7",
  "EPCAM"
)

# Convert gene annotation from the peaks assay into a gene ID to gene symbol map.
gene_map <- as.data.frame(Annotation(atac[["peaks"]])) %>%
  dplyr::select(gene_id, gene_name) %>%
  mutate(gene_id = sub("\\.\\d+$", "", gene_id)) %>%
  distinct(gene_name, .keep_all = TRUE)

# Keep only markers that are present in the ACTIVITY assay.
marker_features <- gene_map %>%
  filter(gene_name %in% lineage_markers) %>%
  filter(gene_id %in% rownames(atac_filtered[["ACTIVITY"]]))

# DotPlot needs Ensembl gene IDs, while plot labels are shown as gene symbols.
features_use <- marker_features$gene_id
names(features_use) <- marker_features$gene_name

feature_labels <- setNames(
  names(features_use),
  unname(features_use)
)

# Visualise whether transferred labels are supported by expected marker activity.
DotPlot(
  atac_filtered,
  features = unname(features_use),
  group.by = "predicted.id"
) +
  scale_x_discrete(labels = feature_labels) +
  RotatedAxis() +
  labs(
    title = "Marker gene activity in transferred ATAC labels",
    x = "Marker gene",
    y = "Transferred cell type"
  )
```
