---
title: "scRNA clustering and annotating"
author:
- name: Zack Guo
  affiliation: UCL
  email: wegmzgu@ucl.ac.uk
date: "`r format(Sys.time(), '%B %e, %Y')`"
output:
  html_document:
    highlight: tango
    toc: yes
    toc_float: yes
    toc_depth: '5'
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Library packages
```{r}
library(Matrix)
library(Seurat)
library(dplyr)
library(biomaRt)
library(openxlsx)
library(ggplot2)
library(readxl)
library(purrr)
library(tidyr)
library(pheatmap)
```

# scRNA clustering to generate 13 clusters and inter-cluster DEGs are found. Parameter and filtering is identical to the paper that generates the scRNA and scATAC data.
```{r, eval=FALSE}
setwd("C:/Bioinfo data")

# Define function to read 10x sample and do QC
load_10x_sample <- function(base_dir, sample_name, sample_label) {
  base_dir <- gsub("\\\\", "/", base_dir)
  mtx <- readMM(gzfile(file.path(base_dir, paste0(sample_name, "_matrix.mtx.gz"))))
  barcodes <- readLines(gzfile(file.path(base_dir, paste0(sample_name, "_barcodes.tsv.gz"))))
  features <- read.table(gzfile(file.path(base_dir, paste0(sample_name, "_features.tsv.gz"))),
                         sep = "\t", header = FALSE, stringsAsFactors = FALSE)
  
  rownames(mtx) <- features$V1
  colnames(mtx) <- barcodes
  
  seu <- CreateSeuratObject(counts = mtx)
  seu$sample <- sample_label
  seu[["gene_symbol"]] <- features$V2
  
# Cell filtering. Cells with low RNA count, low amount of genes, and high mitochondiral gene proportion will be filtered
  seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")
  seu <- subset(seu, subset = nCount_RNA > 1500 & nFeature_RNA > 200 & percent.mt < 50)
  
# Gene filtering. Gene should be present in at least three cells
  counts <- GetAssayData(seu, slot = "counts")
  genes_to_keep <- rowSums(counts > 0) >= 3
  seu <- seu[genes_to_keep, ]
  
  return(seu)
}

# Load four sample
base_dir <- "C:/Bioinfo data/Used before/scRNA"
hm1 <- load_10x_sample(base_dir, "GSM4832483_HM-1", "HM_1")
hm2 <- load_10x_sample(base_dir, "GSM4832484_HM-2", "HM_2")
dm1 <- load_10x_sample(base_dir, "GSM4832481_DM-1", "DM_1")
dm2 <- load_10x_sample(base_dir, "GSM4832482_DM-2", "DM_2")

samples_list <- list(HM_1 = hm1, HM_2 = hm2, DM_1 = dm1, DM_2 = dm2)
samples_list <- lapply(samples_list, function(seu) {
  seu <- NormalizeData(seu)
  seu <- FindVariableFeatures(seu)
  return(seu)
})

# Integration and clustering
anchors <- FindIntegrationAnchors(object.list = samples_list, dims = 1:30)
merged <- IntegrateData(anchorset = anchors, dims = 1:30)

DefaultAssay(merged) <- "integrated"
merged <- ScaleData(merged)
merged <- RunPCA(merged, npcs = 30)
merged <- FindNeighbors(merged, dims = 1:30)
merged <- FindClusters(merged, resolution = 0.6)
merged <- RunUMAP(merged, dims = 1:30)
DimPlot(merged, reduction = "umap", group.by = "seurat_clusters", label = TRUE)

# Find inter-cluster DEGs
DefaultAssay(merged) <- "RNA"
merged <- JoinLayers(merged)
Idents(merged) <- "seurat_clusters"
all_markers <- FindAllMarkers(merged,
                              only.pos = TRUE,
                              min.pct = 0.25,
                              logfc.threshold = 0.25)
```

# Load pre-saved RDS to save time for calculation
```{r}
obj <- readRDS("C:/Bioinfo data/RDS to skip calc/HM_DM_seurat_analysis_with_markers.rds")

merged <- obj$merged
all_markers <- obj$all_markers

Idents(merged) <- "seurat_clusters"

DimPlot(
  merged,
  reduction = "umap",
  group.by = "seurat_clusters",
  label = TRUE,
  repel = TRUE
) + NoLegend()
```

# DEGs:Ensembl to HGNC. Genes without HGNC symbol will be filtered
```{r, eval=FALSE}
# Ensembl to HGNC
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
gene_mapping <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = all_markers$gene,
  mart = mart
)

all_markers_annot <- merge(all_markers, gene_mapping,
                           by.x = "gene", by.y = "ensembl_gene_id", all.x = TRUE)

# Filter gene without HGNC
all_markers_filtered <- all_markers_annot %>%
  filter(!is.na(hgnc_symbol) & hgnc_symbol != "") %>%
  dplyr::distinct(cluster, hgnc_symbol, .keep_all = TRUE)

wb <- createWorkbook()
clusters <- sort(unique(all_markers_filtered$cluster))

for (cl in clusters) {
  sheet_name <- paste0("cluster", cl)
  
  df <- all_markers_filtered %>%
    filter(cluster == cl, avg_log2FC > 0) %>%
    dplyr::select(hgnc_symbol, avg_log2FC, pct.1, pct.2, p_val_adj) %>%
    arrange(desc(avg_log2FC))
  
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, df)
}

saveWorkbook(
  wb,
  "C:/Bioinfo data/cell of origin/Past scripts/cluster_upregulated_by_sheet.xlsx",
  overwrite = TRUE
)
```

# Read excel file for inter-cluster DEG
```{r}
excel_path <- "C:/Bioinfo data/cell of origin/Past scripts/cluster_upregulated_by_sheet.xlsx"

sheet_names <- excel_sheets(excel_path)

markers_by_cluster <- setNames(
  lapply(sheet_names, function(s) {
    read_excel(excel_path, sheet = s)
  }),
  sheet_names
)

names(markers_by_cluster)
```
#Calculate overlap
```{r}
# paths
my_excel <- "C:/Bioinfo data/cell of origin/Past scripts/cluster_upregulated_by_sheet.xlsx"

# Reference path
paper_excel <- "C:/Bioinfo data/cell of origin/mmc2.xlsx"

# 1. read my markers
my_sheets <- excel_sheets(my_excel)

my_markers <- map_dfr(my_sheets, function(s) {
  read_excel(my_excel, sheet = s) %>%
    dplyr::select(hgnc_symbol, avg_log2FC) %>%
    mutate(my_cluster = s)
})

# 2. read paper markers
paper_sheets <- excel_sheets(paper_excel)

paper_markers <- map_dfr(paper_sheets, function(s) {
  read_excel(paper_excel, sheet = s, col_names = FALSE) %>%
    dplyr::select(
      gene = 1,
      avg_logFC = 3
    ) %>%
    mutate(ref_cluster = s)
})

# 3. clean genes and rank only by fold change
top_n <- 100

my_top <- my_markers %>%
  filter(!is.na(hgnc_symbol), hgnc_symbol != "") %>%
  group_by(my_cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE) %>%
  slice_head(n = top_n) %>%
  summarise(genes = list(unique(hgnc_symbol)), .groups = "drop")

paper_top <- paper_markers %>%
  filter(!is.na(gene), gene != "") %>%
  group_by(ref_cluster) %>%
  arrange(desc(avg_logFC), .by_group = TRUE) %>%
  slice_head(n = top_n) %>%
  summarise(genes = list(unique(gene)), .groups = "drop")

# 4. calculate overlap
score_df <- expand.grid(
  my_cluster = my_top$my_cluster,
  ref_cluster = paper_top$ref_cluster,
  stringsAsFactors = FALSE
) %>%
  rowwise() %>%
  mutate(
    my_genes = list(my_top$genes[[match(my_cluster, my_top$my_cluster)]]),
    ref_genes = list(paper_top$genes[[match(ref_cluster, paper_top$ref_cluster)]]),
    overlap_n = length(intersect(my_genes, ref_genes)),
    overlap_genes = paste(intersect(my_genes, ref_genes), collapse = ", ")
  ) %>%
  ungroup() %>%
  dplyr::select(my_cluster, ref_cluster, overlap_n, overlap_genes)

# 5. best match for each of my clusters, ordered by cluster number

get_cluster_num <- function(x) {
  as.numeric(gsub("[^0-9]", "", x))
}

best_match <- score_df %>%
  mutate(
    my_cluster_num = get_cluster_num(my_cluster),
    ref_cluster_num = get_cluster_num(ref_cluster)
  ) %>%
  group_by(my_cluster) %>%
  arrange(desc(overlap_n), ref_cluster_num, .by_group = TRUE) %>%
  slice_head(n = 3) %>%
  ungroup() %>%
  arrange(my_cluster_num, desc(overlap_n)) %>%
  dplyr::select(my_cluster, ref_cluster, overlap_n, overlap_genes)

best_match
```

# Heatmap ordered by cluster number
```{r marker-overlap-heatmap, fig.width=8, fig.height=6}
get_cluster_num <- function(x) {
  as.numeric(gsub("[^0-9]", "", x))
}

my_cluster_order <- score_df %>%
  distinct(my_cluster) %>%
  mutate(my_cluster_num = get_cluster_num(my_cluster)) %>%
  arrange(my_cluster_num) %>%
  pull(my_cluster)

ref_cluster_order <- score_df %>%
  distinct(ref_cluster) %>%
  mutate(ref_cluster_num = get_cluster_num(ref_cluster)) %>%
  arrange(ref_cluster_num) %>%
  pull(ref_cluster)

score_mat <- score_df %>%
  mutate(
    my_cluster = factor(my_cluster, levels = my_cluster_order),
    ref_cluster = factor(ref_cluster, levels = ref_cluster_order)
  ) %>%
  dplyr::select(my_cluster, ref_cluster, overlap_n) %>%
  pivot_wider(
    names_from = ref_cluster,
    values_from = overlap_n,
    values_fill = 0
  ) %>%
  arrange(my_cluster) %>%
  as.data.frame()

# Y axis = your clusters
rownames(score_mat) <- paste0("My_", score_mat$my_cluster)
score_mat$my_cluster <- NULL

# reorder columns
score_mat <- score_mat[, ref_cluster_order, drop = FALSE]

# X axis = paper reference clusters
colnames(score_mat) <- paste0("P_", get_cluster_num(colnames(score_mat)))


pheatmap(
  as.matrix(score_mat),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
  angle_col = 45,
  main = "Marker overlap\nRows = My clusters | Columns = Paper reference clusters"
)
```
# Make a annotation table. Reference: Figure 1D
```{r}
annotation_table <- data.frame(
  cluster = paste0("cluster", 0:12),
  paper_cluster = c(
    "P0",
    "P2",
    "P8",
    "P6",
    "P3",
    "P7",
    "P1",
    "P1",
    "P9",
    "P4",
    "P10",
    "P12",
    "P11"
  ),
  cell_type = c(
    "Mature hepatocytes 1",
    "Mesenchymal cells 1",
    "Bi-potent progenitor",
    "Cholangiocytes 2",
    "Hepatocyte-like 2",
    "Mesenchymal cells 2",
    "PHH-like cells",
    "PHH-like cells",
    "Proliferating cells",
    "Hepatocyte-like 1",
    "Mature hepatocytes 2",
    "Hepatic stellate cells",
    "Immune cells"
  )
)

annotation_table
```

# Identity check only for billary cells and hepatocytes
```{r}
# Only genes present in Seurat object
markers_check_id <- c(
  "ENSG00000171557", # FGG
  "ENSG00000172716", # ACSM3
  "ENSG00000130208", # APOC1
  "ENSG00000106258", # CYP3A5
  "ENSG00000160182", # TFF1
  "ENSG00000171747", # KRT7
  "ENSG00000117318"  # EPCAM
)

# Rename x-axis labels
feature_labels <- c(
  "ENSG00000171557" = "FGG",
  "ENSG00000172716" = "ACSM3",
  "ENSG00000130208" = "APOC1",
  "ENSG00000106258" = "CYP3A5",
  "ENSG00000160182" = "TFF1",
  "ENSG00000171747" = "KRT7",
  "ENSG00000117318" = "EPCAM"
)

DotPlot(
  merged,
  features = markers_check_id,
  group.by = "seurat_clusters"
) +
  scale_x_discrete(labels = feature_labels) +
  RotatedAxis()

#Hepatocyte identity was supported by the expression of hepatocyte-associated markers (FGG, APOC1 and CYP3A5), particularly in hepatocyte-like clusters. Biliary/progenitor populations showed higher expression of epithelial markers (TFF1 and EPCAM), supporting their annotation as cholangiocyte- or progenitor-like populations.
```
# Rremove uncertain cluster and save RDS
```{r}
# Define ambiguous clusters to remove before ATAC label transfer
clusters_remove <- c("1", "5", "8", "9", "12")

# Use Seurat clusters as active identities
Idents(merged) <- "seurat_clusters"

# Remove ambiguous clusters from the scRNA reference object
merged_for_atac <- subset(
  merged,
  idents = clusters_remove,
  invert = TRUE
)

# Remove unused cluster levels after subsetting
merged_for_atac$seurat_clusters <- droplevels(
  factor(merged_for_atac$seurat_clusters)
)

# Keep only annotations for the remaining clusters
annotation_table_filtered <- annotation_table %>%
  filter(!(gsub("cluster", "", cluster) %in% clusters_remove))

# Keep only annotations for the remaining clusters
annotation_table_filtered <- annotation_table %>%
  filter(!(gsub("cluster", "", cluster) %in% clusters_remove))

# Create cluster-to-cell-type map
celltype_map <- annotation_table_filtered$cell_type
names(celltype_map) <- gsub("cluster", "", annotation_table_filtered$cluster)

# Add cell type annotation to the filtered scRNA object
merged_for_atac$cell_type <- unname(
  celltype_map[as.character(merged_for_atac$seurat_clusters)]
)

# Check annotation
table(merged_for_atac$seurat_clusters)
table(merged_for_atac$cell_type)
table(is.na(merged_for_atac$cell_type))

# Save one filtered scRNA reference object for ATAC integration
saveRDS(
  merged_for_atac,
  "C:/Bioinfo data/RDS to skip calc/HM_DM_scRNA_reference_for_ATAC.rds"
)
```

